Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R

### 2017-04-02 rmark_doc_inla_nmixnb_8.Rmd
More updates

---
title: Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R
author: Timothy D. Meehan^[National Audubon Society, Boulder, Colorado USA], Nicole L. Michel^[National Audubon Society, San Francisco, California USA], and Haavard Rue^[King Abdulla University of Science and Technology, Thuwal SAU]
date: "March 31, 2017"
output: pdf_document
---

@@

```{r make plot 1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 1. Marginal posteriors from JAGS (solid gray lines) and R-INLA (dashed black lines)."}
# get output from jags
jags.mcmc <- combine.mcmc(out.jags, thin=3, return.samples=5000)
jags.df <- as.data.frame(jags.mcmc)
# make density for intLam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[1] for NMix observations`)
d1$x <- exp(d1$x)
d2 <- as.data.frame(density(jags.df$intLam)[c(1,2)])
d2$x <- exp(d2$x)
p1 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Lambda") + ylab(" ") + scale_y_continuous(breaks=c(3,6,9)) + theme_acbs() + geom_vline(xintercept=exp(2), col="gray20", lty=1)
# make density for x.1.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[2] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bCov1Lam)[c(1,2)])
p2 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 1")  + ylab("Posterior density") + theme_acbs() + geom_vline(xintercept=2, col="gray20", lty=1)
# make density for x.2.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[3] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bCov2Lam)[c(1,2)])
p3 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 2")  + ylab(" ") + theme_acbs() + geom_vline(xintercept=-3, col="gray20", lty=1)
# make density for yr.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[4] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bYr)[c(1,2)])
p4 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Year covariate")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)
# make density for intP
d1 <- as.data.frame(out.inla$marginals.fixed$`(Intercept)`)
d1$x <- plogis(d1$x)
d2 <- as.data.frame(density(jags.df$intP)[c(1,2)])
d2$x <- plogis(d2$x)
p5 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("p")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=plogis(1), col="gray20", lty=1)
# make density for x.1.p
d1 <- as.data.frame(out.inla$marginals.fixed$x.p.1.inla)
d2 <- as.data.frame(density(jags.df$bCov1P)[c(1,2)])
p6 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 1")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=-2, col="gray20", lty=1)
# make density for x.3.p
d1 <- as.data.frame(out.inla$marginals.fixed$x.p.3.inla)
d2 <- as.data.frame(density(jags.df$bCov3P)[c(1,2)])
p7 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 3")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)
# make density for overdisp
d1 <- as.data.frame(out.inla$marginals.hyperpar$`overdispersion for NMix observations`)
d1$x <- 1 / d1$x; d1$y <- d1$y / 8
d2 <- as.data.frame(density(jags.df$overDisEst)[c(1,2)])
p8 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Overdisperison parameter") + ylab(NULL) + theme_acbs() + geom_vline(xintercept=3, col="gray20", lty=1)
# plot all
multiplot(p1,p2,p3,p4,p5,p6,p7,p8,cols=3)
```

@@

```{r explore sim results, echo=FALSE}
# explore results
par_names <- c(
  'bCov1Lam'="Abundance covariate 1 (2)",
  'bCov2Lam'="Abundance covariate 2 (-3)",
  'bYr'="Abundance covariate year (1)",
  'bCov1P'="Detection covariate 1 (-2)",
  'bCov3P'="Detection covariate 3 (X)",
  'intLam'="Abundance intercept (2)",
  'intP'="Detection intercept (1)",
  'overDisEst'="Overdispersion estimate (3)"
)
to.string <- as_labeller(par_names)
ggplot(data=sim.out, aes(x=p.b3, y=diffs, colour=simtype, linetype=simtype)) + geom_point(pch=1, size=0.5) +
  facet_wrap(~factor(par), scales="free",labeller=to.string) +
  xlab("Slope coefficient for the detection covariate") +
  ylab("Difference between posterior mean and true parameter value") +
  geom_smooth(span=3, se=F, size=0.5) + 
  theme_acbs() + scale_color_manual(values=c("black","gray50")) + 
  theme(strip.background = element_rect(colour = "black", fill = "white"), 
        legend.position="none") + scale_linetype_manual(values=c(1,2))
```

>>>>

---
title: Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R
author: Timothy D. Meehan^[National Audubon Society, Boulder, Colorado USA], Nicole L. Michel^[National Audubon Society, San Francisco, California USA], and Haavard Rue^[King Abdulla University of Science and Technology, Thuwal SAU]
date: "March 31, 2017"
output: 
  pdf_document: 
    fig_caption: yes
---

@@

```{r make plot 1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 1. Marginal posteriors from JAGS (solid gray lines) and R-INLA (dashed black lines)."}
# get output from jags
jags.mcmc <- combine.mcmc(out.jags, thin=3, return.samples=5000)
jags.df <- as.data.frame(jags.mcmc)
# make density for intLam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[1] for NMix observations`)
d1$x <- exp(d1$x)
d2 <- as.data.frame(density(jags.df$intLam)[c(1,2)])
d2$x <- exp(d2$x)
col1 <- 4
d3 <- all.out.tab[col1, 5]
d4 <- all.out.tab[col1, 4]
d5 <- all.out.tab[col1, 6]
d6 <- exp(c(d3,d4,d5))
d7 <- data.frame(y=c(0,0,0), x=d6)
p1 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Lambda") + ylab(" ") + scale_y_continuous(breaks=c(3,6,9)) + theme_acbs() + geom_vline(xintercept=exp(2), col="gray20", lty=1) + 
geom_line(data=d7, aes(y=y, x=x)) +
geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
# make density for x.1.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[2] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bCov1Lam)[c(1,2)])
col1 <- 5
d3 <- all.out.tab[col1, 5]
d4 <- all.out.tab[col1, 4]
d5 <- all.out.tab[col1, 6]
d6 <- c(d3,d4,d5)
d7 <- data.frame(y=c(0,0,0), x=d6)
p2 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 1")  + ylab("Posterior density") + theme_acbs() + geom_vline(xintercept=2, col="gray20", lty=1) + 
geom_line(data=d7, aes(y=y, x=x)) +
geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
# make density for x.2.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[3] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bCov2Lam)[c(1,2)])
col1 <- 6
d3 <- all.out.tab[col1, 5]
d4 <- all.out.tab[col1, 4]
d5 <- all.out.tab[col1, 6]
d6 <- c(d3,d4,d5)
d7 <- data.frame(y=c(0,0,0), x=d6)
p3 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 2")  + ylab(" ") + theme_acbs() + 
geom_vline(xintercept=-3, col="gray20", lty=1) + 
geom_line(data=d7, aes(y=y, x=x)) +
geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
# make density for yr.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[4] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bYr)[c(1,2)])
col1 <- 7
d3 <- all.out.tab[col1, 5]
d4 <- all.out.tab[col1, 4]
d5 <- all.out.tab[col1, 6]
d6 <- c(d3,d4,d5)
d7 <- data.frame(y=c(0,0,0), x=d6)
p4 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Year covariate")  + ylab(NULL) + theme_acbs() + 
geom_vline(xintercept=1, col="gray20", lty=1) + 
geom_line(data=d7, aes(y=y, x=x)) +
geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
# make density for intP
d1 <- as.data.frame(out.inla$marginals.fixed$`(Intercept)`)
d1$x <- plogis(d1$x)
d2 <- as.data.frame(density(jags.df$intP)[c(1,2)])
d2$x <- plogis(d2$x)
p5 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("p")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=plogis(1), col="gray20", lty=1)
# make density for x.1.p
d1 <- as.data.frame(out.inla$marginals.fixed$x.p.1.inla)
d2 <- as.data.frame(density(jags.df$bCov1P)[c(1,2)])
p6 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 1")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=-2, col="gray20", lty=1)
# make density for x.3.p
d1 <- as.data.frame(out.inla$marginals.fixed$x.p.3.inla)
d2 <- as.data.frame(density(jags.df$bCov3P)[c(1,2)])
p7 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 3")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)
# make density for overdisp
d1 <- as.data.frame(out.inla$marginals.hyperpar$`overdispersion for NMix observations`)
d1$x <- 1 / d1$x; d1$y <- d1$y / 8
d2 <- as.data.frame(density(jags.df$overDisEst)[c(1,2)])
p8 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Overdisperison parameter") + ylab(NULL) + theme_acbs() + geom_vline(xintercept=3, col="gray20", lty=1)
# plot all
multiplot(p1,p2,p3,p4,p5,p6,p7,p8,cols=3)

@@

```{r explore sim results, echo=FALSE}
# explore results
par_names <- c(
  'bCov1Lam'="Abundance covariate 1 (2)",
  'bCov2Lam'="Abundance covariate 2 (-3)",
  'bYr'="Abundance covariate year (1)",
  'bCov1P'="Detection covariate 1 (-2)",
  'bCov3P'="Detection covariate 3 (X)",
  'intLam'="Abundance intercept (2)",
  'intP'="Detection intercept (1)",
  'overDisEst'="Overdispersion estimate (3)"
)
to.string <- as_labeller(par_names)
ggplot(data=sim.out, aes(x=p.b3, y=diffs, colour=simtype, linetype=simtype)) + geom_point(pch=1, size=0.5) +
  facet_wrap(~factor(par), scales="fixed",labeller=to.string) +
  xlab("Slope coefficient for the detection covariate") +
  ylab("Difference between posterior mean and true parameter value") +
  geom_smooth(span=3, se=F, size=0.5) + 
  theme_acbs() + scale_color_manual(values=c("black","gray50")) + 
  theme(strip.background = element_rect(colour = "black", fill = "white"), 
        legend.position="none") + scale_linetype_manual(values=c(1,2))
```
